---
title: "Interactive Maps: COVID-19 Testing Locations using Leaflet"
author: "Sunder Sai"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyr)
library(tidyverse)
library(broom)
library(lubridate)
library(janitor)
library(knitr)
```

***
#### In this project, I use data to clean and plot **COVID-19 Testing Locations.** Using the `leaflet` package, I make these maps interactive with zoom capabilities and pop-up displays with information specific to each datapoint on the map. I also show how to cluster data points together and swap map backgrounds. These interactive maps can serve as an important resource for those seeking information regarding testing locations near them.       
***  
  
### 1.1) Reading in data and cleaning  

* Testing location data has been made publicly available by OscarHealth on their Github website found [here](https://github.com/oscarhealth/covid19-testing)      
* I will first read the dataset in, and then clean the dataset by removing unwanted columns such as those that include missing data      
* This includes removing the following columns: `visible`, `facility_id`, `phone`, `facility_display_note`, and those that start with `hours_of_operation`    

```{r}

testing_location <- read.csv("data/testing_locations.csv")
testing_location <- select(testing_location, -starts_with("hours_of_operation"))
testing_location$visible <- NULL
testing_location$facility_id <- NULL
testing_location$phone <- NULL
testing_location$facility_display_note <- NULL

kable(head(testing_location))
```

### 1.2) Creating Full Address and Filtering for NYState     
* Next, I notice that the addresses are split up  by street, zipcode, and state. I would like a variable with these aspects consolidated in one column    
* We can do this using the `paste` function and separating each portion with comma    
```{r}
testing_location$full_address <- paste(testing_location$address, 
                                       testing_location$city, 
                                       testing_location$state,
                                       testing_location$zip, 
                                       sep = ", ")
head(testing_location$full_address)
```

  
  
* I then want to rename the latitude and longitude variables    
* Next, I want to filter for New York State and New York City    
```{r}
# renaming latitude = lat, longitude = lng 
testing_location <- testing_location %>% 
  rename(latitude = lat, 
         longitude = lng)

# filtering for new york
ny_testinglocation <- testing_location %>%
  filter(state == "NY")

# filtering for new york city 
nyc_testinglocation <- ny_testinglocation %>%
  filter(city == "Bronx"|
           city == "New York"|
           city == "Queens"|
           city == "Brooklyn"|
           city == "Staten Island")

kable(head(nyc_testinglocation))
```


### 1.3) Plotting basic points and exploring leaflet  
* Below, I plot the New York City testing points with no background to see the distribution of points      

```{r}
require(leaflet)

# graphing with no background: nyc
ggplot(data = nyc_testinglocation) +
  geom_point(aes(x = longitude, y = latitude), alpha = 0.2) +
  theme_bw() 
```

* I also check out the leaflet package before adding in the points.   
* Using `addTiles()`, I set a generic background and then through `setView` I can set the map frame to focus in on New York 
```{r}
# basic leaflet background 
require(leaflet)
leaflet() %>%
  addTiles() %>%
  setView(-74, 40.755, zoom = 11.5)
```


### 1.4) Creating Interactive Maps with Leaflet    

* Below, I create an interactive map with points set to the latitude and longitude of the testing data      
* I also include the `popup` argument that provides the relevant information for the Facility Name and the full Address details    
```{r}
map <- leaflet(data=ny_testinglocation) %>%
  addTiles() %>%
  addMarkers(
    lng = ~longitude,
    lat = ~latitude, 
    popup = paste("Facility Name:", nyc_testinglocation$facility_name, "<br>",
                  "Address:", nyc_testinglocation$full_address))

map
```

* This map is **interactive** so feel free to zoom in out.   
* Click on each point to see the relevant information, including **Facility Name** and **Address info.** Pretty cool!  


### 1.5) Adding Cluster Points and Changing Backgrounds      

* The points, however, may look a little crowded together      
* We can cluster nearby points together into spatial clusters to make it more visually appealing    
* This can be done through the `clusterOptions = markerClusterOptions()` argument     
```{r}

map1 <- leaflet(data=ny_testinglocation) %>%
  addTiles() %>%
  addMarkers(
    lng = ~longitude,
    lat = ~latitude, 
    popup = paste("Facility Name:", nyc_testinglocation$facility_name, "<br>",
                  "Address:", nyc_testinglocation$full_address),
    clusterOptions = markerClusterOptions())

map1
```
**Zooming** into the map can dive into the clusters of points, revealing further sub-clusters. This is a great way to visualize clustering of testing locations on the interactive map.  

**Backgrounds:**  
*  We can also change the backgrounds by passing our `map` object through the `addProviderTiles()` function      
```{r}
map %>% addProviderTiles(providers$Esri.NatGeoWorldMap)

```

*  Above is Esri's Nat-Geo World Map background   

   
```{r}
map %>% addProviderTiles(providers$CartoDB.Positron)
```

*  Here is the positron and darkmatter background 

   
```{r}
map %>% addProviderTiles(providers$Stamen.Toner)
```
*  Above is the black and white toner background 

   
```{r}
map %>% addProviderTiles(providers$MtbMap)
```

*  Here, is the MTB mountain bike and trail map 