---
title: "Mapping SAT Scores Across the US"
author: "Sunder Sai"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

## Introduction: 
#### In this project I will be working with Spatial Packages in R to access, manage, and export spatial data for visualization. I use SAT Math and Verbal test scores that are publicly available from College Board on their [website](https://collegereadiness.collegeboard.org/educators/k-12/educator-downloads). I completed this project as a graduate student at Columbia University Mailman School of Public Health as part of the Advanced GIS & Spatial Analysis course under the guidance of Dr. Jeremy Porter. 

***
## **Part 1: Projections**
***

### 1.1) The World Boundaries
* Here, I load required packages, assign the world boundary file to the object `world` and then examine the units by displaying country names
```{r, warning=FALSE}
library(maps) 

oldpar<-par()

world <- map("world", res=0)
str(world)
head(world$names)
plot(world)
```




### 1.2) US Boundaries
* Next, I assign the state boundary file to the object `states` and display names of states
```{r}
states <- map("state", res=0)
str(states)
head(states$names)
plot(states)
```

### 1.3) Spatial Lines

* Using `maptools` I create spatial lines objects using long/lat projections
```{r, warning=FALSE}
library(maptools)

spworld <- map2SpatialLines(world, proj4string = CRS("+proj=longlat"))
spstates <- map2SpatialLines(states, proj4string = CRS("+proj=longlat"))

str(spworld, max.level=2)
str(spstates,max.level=2)

plot(spworld)
plot(spstates)
```

### 1.4) Long/Lat Projections

* Using the `rgdal` package we can examine differences in displays of long/lat projections.
```{r, warning=FALSE}
library(rgdal)

world.laea <- spTransform(spworld, CRS("+proj=laea +lat_0=0 +lon_0=0"))
states.laea <- spTransform(spstates, CRS("+proj=laea +lat_0=43.0758 +lon_0=-89.3976"))
states.epsg <- spTransform(spstates, CRS("+init=epsg:3623"))
```

* This appears better when we use the Lambert Azmuthal Equal Area (known as laea) projection 
```{r, warning=FALSE}

#Run Following Code chunk together


par(mfrow = c(2, 2), pty = "s", cex.axis = 0.5)

plot(spworld, axes = T)
title(main = "Longitude and\nLatitude")
plot(world.laea, axes = T)
title(main = "Lambert Azimuthal\nEqual Area")
plot(spstates, axes = T)
title(main = "Longitude and\nLatitude")
plot(states.laea, axes = T)
title(main = "Lambert Azimuthal\nEqual Area")
```

***
## **Part 2: Spatial Referencing**
***


### 2.1: Linking Data to Spatial Objects in R

* Here, I create the `map.states` object by calling the states boundaries.
* I then list the state names by using `strsplit` function to separate their true names from other location aspects
```{r, warning=FALSE}

par(oldpar)

map.states <- map("state", plot = T, fill = T, res=0)
str(map.states)

names(map.states)
head(map.states$names)

list.names.states <- strsplit(map.states$names,":")
head(list.names.states)
```


### 2.2) Converting to Spatial Polygons

* I then convert the `map.states` object to a Spatial Polygon with IDs set to `map.IDs`
* I visualize our base map here, below
```{r}
map.IDs <- sapply(list.names.states, function(x) x[1])
head(map.IDs, n=63)

polystates <- map2SpatialPolygons(map.states, IDs = map.IDs,proj4string = CRS("+proj=longlat"))

summary(polystates)

plot(polystates)
```

### 2.3) Projecting States

* Here, I project the `states` object using the laea projection command
* I then set and list the spatial polygon IDs for `states.laea`
* This map is more accurate in terms of curvature of the globe
```{r}
states.laea <- spTransform(polystates, CRS("+proj=laea +lat_0=43.0758 +lon_0=-89.3976"))
plot(states.laea)

sp.IDs <- sapply(slot(states.laea, "polygons"), function(x) slot(x,"ID"))
head(sp.IDs, n=50)
```

***
## **Part 3: SAT Data**  
***

### 3.1) Downloading data and creating spatial dataframe 

* First, I download the data and view its contents
* There are 50 states, its abbreviations, and the Math scores
```{r}
#START
sat_math<- read.csv("data/sat_math.csv", stringsAsFactors = F,row.names = 1)
head(sat_math, n=50)
```

* Then, I create a spatial dataframe
* I plot a base laea version of this here below
```{r}
states.math <- SpatialPolygonsDataFrame(polystates,sat_math)
summary(states.math)

states.math.laea <- spTransform(states.math, CRS("+proj=laea +lat_0=43.0758 +lon_0=-89.3976"))

plot(states.math.laea)
```

* We can view a brief summary of this below
```{r}
summary(states.math.laea)
```

### 3.2) Color Ramps for Mapping

* We need to first install and load the color brewer package
* Then we can visualize the different color ramps as seen below
```{r}
library(RColorBrewer)
display.brewer.all()
display.brewer.pal(5, "Greys")
```

* These different color ramps can be used to represent different variations in SAT scores

***
## **Part 4) Mapping**
***

### 4.1) Visualize distribution of SAT math scores

* I use the `classInt` package to set class categories 
* Using the following code, I visualize the distribution of SAT verbal Scores broken into Quantiles

```{r, warning=FALSE}
library(classInt)

plotvar <- states.math.laea$math
nclr <- 5
plotclr <- brewer.pal(nclr, "Greys")
plotclr
class <- classIntervals(plotvar, nclr, style = "quantile")
class
```

* Here we can check out the hex codes for the different colors 
* Below is also a black and white map
```{r, warning=FALSE}
colcode <- findColours(class, plotclr, digits = 3)
colcode
plot(states.math.laea, col = colcode)
```

### 4.2) Adding color, legends, and titles

* Finally, we can set a color scheme of "blues", a `style` of "quantiles" as well as a title and legend 

```{r}
plotclr <- brewer.pal(nclr, "Blues")
class <- classIntervals(plotvar, nclr, style = "quantile")
colcode <- findColours(class, plotclr, digits = 3)
plot(states.math.laea, col = colcode, border = "grey",axes = F)
title(main = "SAT Math Scores in 2010 \n by Sunder Sai", 
      sub = "Data Source: College Board")
legend("bottomleft", legend = names(attr(colcode,"table")), 
       fill = attr(colcode, "palette"), cex=0.45)

```


